---
title: Time Series Analysis of average yearly salary of individual Within Canada and
  inflation in canada
author: "kush manek"
date: '`r Sys.Date()`'
output:
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(forecast)
library(tseries)
library(astsa)
library(ggplot2)
data <- read.csv("./data.csv");
data$time <- as.Date(data$time)
data <- data[data$yearly_salary > 0 & data$inflation_values > 0, ]
ys <- data$yearly_salary;
ir <- data$inflation_values;
ir <- ts(ir, frequency = 12, start = c(2005, 12))
ys<- ts(ys, frequency = 12, start = c(2005, 12))
plot.ts(ir, main="Inflation Rate Time Series", ylab="Inflation rate", xlab= "Time");
plot.ts(ys, main="Average Yearly Salary Time Series", ylab="Average Yearly salary", xlab= "Time");
acf2(ys,50,main="Yearly Salary ACF and PACF");
acf2(ir,50,main="Inflation Rate ACF and PACF");
```

***Transformations***

To ensure stationarity, the logarithm of the yearly salary was computed:

```{r}
lys <- log(ys);
plot.ts(lys, main="Log Average Yearly Salary Time Series", ylab="Log Average Yearly salary", xlab= "Time");
acf2(lys,50,main="Log Yearly Salary ACF and PACF")
dlys <- diff(lys,12);
plot.ts(dlys, main="Differenced Log Average Yearly Salary Time Series", ylab="Differenced Log Average Yearly salary", xlab= "Time");
acf2(dlys,50,main="Differenced Log Yearly Salary ACF and PACF")
lir <- log(ir);
plot.ts(lir, main="Log Inflation Rate Time Series", ylab="Log Inflation rate", xlab= "Time");
acf2(lir,50,main="Log Inflation Rate ACF and PACF")
dlir <- diff(lir,12);
plot.ts(dlir, main="Differenced Log Inflation Rate Time Series", ylab="DifferencedLog Inflation rate", xlab= "Time");
acf2(dlir,50,main="Differenced Log Inflation Rate ACF and PACF")
```

***Correlation***

```{r}
correlation <- cor(ys, ir, use = "complete.obs", method = "pearson")
print(paste("Pearson Correlation: ", correlation))
ccf(dlys, dlir, main = "CCF: Differenced Log Yearly Salary vs Inflation Rate")
```

***Forecasting***

Inflation rate data was similarly assessed for stationarity. Differencing was applied to stabilize the variance.

Model Fitting

ARIMA Models:

Using ACF and PACF plots, ARIMA models were fitted to both time series:

Inflation Rate Model:

```{r}
ys_optimum  <- sarima(ys,1,1,0,3,1,0,12);
ys_optimum$ttable;
```

```{r}
ir_optimum  <- sarima(ir,2,1,3,2,1,0,12);
ir_optimum$ttable
```
Forecast for the next 12 observations (months) were generated for both variables:

Inflation Rate:

```{r}
sarima.for(ys,1,1,0,3,1,0,12, n.ahead = 12)
```

```{r}
sarima.for(ir,2,1,3,2,1,0,12,n.ahead=12);
```

